---
title: "EDA Planning"
format: html
editor: visual
---

# Loading Library

```{r}
library(tidyverse) 
library(here)
```

# Reading Data

```{r}
data <- read.csv(here("data","final_analytical_data.csv"), row.names = NULL)
```

# Step 1 - Get a quick sense of the data.

-   Look at the **top** of the data

```{r}
head(data) # First 6 rows of data
```

-   Look at the **bottom** of the data

```{r}
tail(data) # Last 6 rows of data
```

-   Look at the **random selection** of the data

```{r}
set.seed(7001)
data %>% slice_sample(n = 6)
```

# Step 2 - Understand the properties of individual variables

-   Variables and corresponding data type

```{r}
glimpse(data)
```

-   The key summary statistics of variables

```{r}
summary(data)
```

# Conclusion from Step 1 & 2:

The data seems fine, but for variables (`armed.conflict`,`OECD`,`OECD2023`,`Drought`, `Earthquake`), these I would code them as factor, since they are binary variables.

And we need aware there are many missing values in the morality rate variables(`Infant.mortality.rate`,`Under.5.mortality.rate`,`Neonatal.mortality.rate`, `Maternal.mortality.rate`), armed conflict variables (`armed.conflict`) and disaster variables (`Drought`,`Earthquake`)

```{r}
data = data %>% mutate(armed.conflict = as.factor(armed.conflict), OECD = as.factor(OECD), OECD2023 = as.factor(OECD2023), Drought = as.factor(Drought), Earthquake = as.factor(Earthquake) )
```

# Step 3 - Understand the distribution of the individual variables

-   For outcome variables (`Infant.mortality.rate`,`Under.5.mortality.rate`,`Neonatal.mortality.rate`, `Maternal.mortality.rate`)

```{r}
data %>% 
  ggplot(aes(x = Infant.mortality.rate)) +
  geom_histogram(bins = 100)
data %>% 
  ggplot(aes(x = Under.5.mortality.rate)) +
  geom_histogram(bins = 100)
data %>% 
  ggplot(aes(x = Neonatal.mortality.rate)) +
  geom_histogram(bins = 100)
data %>% 
  ggplot(aes(x = Maternal.mortality.rate)) +
  geom_histogram(bins = 100)
```

-   Summary: We can see the mortality rate variables are highly right skewed
